### 2013-02-06 Spatial_adaptation/local-adaptation.tex
/* Local Adaptation */

Let $p(t,x)$ be the proportion of the allele under spatially varying selection $s(x)$ in one dimension,
so that 
\[
    \partial_t p(t,x) = \sigma^2 \partial_x^2 p(t,x) + s(x) p(t,x) (1-p(t,x)) .
\]
The stable distribution $\phi(x) = \lim_{t\to\infty} p(t,x)$ then solves
\begin{align} \label{eqn:definingphi}
    \partial_x^2 \phi(x) = - s(x) \phi(x) (1-\phi(x)) /sigma^2,
\end{align}
with appropriate boundary conditions.
First rescale space by $\sigma$ so that it dissappears.
If we now assume that $s(x)$ is piecewise constant,
$s(x) = s_i$ for $x \in [x_i,x_{i+1})$, with $x_0=-\infty$ and $x_{n+2}=\infty$,
then the equation is integrable: if we multiply through by $\partial_x \phi(x)$ and integrate, then we get that
\begin{align} \label{eqn:conservation}
    ( \partial_x \phi(x) )^2  &= - \int^{x} s(x) \phi(x) (1-\phi(x)) \partial_x \phi(x) dx \\
        &= - \frac{1}{2} s_i \phi(x)^2 \left( 1 - \frac{1}{3} \phi(x) \right) - K_i \quad \mbox{for } x \in [x_i,x_{i+1}) \\
        &:= - V_i(\phi(x)) \quad \mbox{for } x \in [x_i,x_{i+1}) ,
\end{align}
where $K_i$ is a constant of integration and we define
\[
        V_i(\phi) = \frac{1}{2} s_i \phi^2 \left( 1 - \frac{1}{3} \phi \right) + K_i .
\]
Note that $V_i'(0)=V_i'(1)=0$, that $V_i(0)=K_i$ and $V_i(1) = K_i-s_i/6$.  
We will always have that $V(\phi) \le 0$.
(We have then that $\phi'' = - \partial_\phi V(\phi)$, the equation of motion of a particle in potential $V$.
Also note that this implies ``conservation of energy'', i.e.\ $( \partial_x \phi(x) )^2 + V(\phi(x))$ is constant.)
Rearranging, we get that $dx = d\phi / \sqrt{-V(\phi)}$, so where $\phi(x)$ is monotone, the inverse is
\[
    x(\phi) = x(\phi_0) \pm \int_{\phi_0}^\phi \frac{ d\psi }{ \sqrt{ -V(\psi) } } .
\]
For each $i$ then define the elliptic function
\[
    F_i(\phi) = \int_{\phi_i^*}^\phi \frac{ d\psi }{ \sqrt{ -V(\psi) } } ,
\]
where take the positive branch of the square root, and $\phi_i^*$ will be chosen later.
Then we have that
$x(\phi) - x(\phi_0) = \pm( F(\phi) - F(\phi_0))$,
or for an appropriate $x_0$,
\[
    \phi(x) = F^{-1}\left( F(\phi(x_0)) \pm (x - x_0) \right).
\]

@@

We also require that $\phi(x)$ and $\phi'(x)$ are continuous.
Continuity of $\phi'(x)$ is equivalent to $V_i(x_{i+1}) = V_{i+1}(x_{i+1})$,
which we can rearrange to find an equation for $K_{i+1}$ in terms of $K_i$ and $\phi(x_{i+1})$:
\[
    K_{i+1} - K_i = (s_i - s_{i+1}) \frac{1}{2} \phi(x_{i+1})^2 (1-\phi(x_{i+1})/3) .
\]
What about $\phi(x_i)$?  Well, if $\phi(x)$ is monotone on $[x_i,x_{i+1})$ then we can wolog take $\phi_i^*=0$ or $1$ depending on the sign of $s_i$.
Otherwise, let $\phi_i^*$ be the (unique) root of $V_i$ in $[x_i,x_{i+1})$, so $V_i(\phi_i^*)=0$.
In this case, $\phi_i^*$ is the maximum or minimum of $\phi$ in the interval:
if $s>0$, then $\phi_i = \max\{ \phi(x) : x \in [x_i,x_{i+1})\}$. 
Recall we defined $F_i$ using $\phi_i^*$; now using the fact that $\phi$ is monotone with the opposite sign on either side of $\phi_i^*$,
$x_{i+1} - x_i = F(\phi(x_{i+1})) - F(\phi_i^*) + F(\phi(x_{i})) - F(\phi_i^*)$,
and that $F(\phi_i^*) = 0$,
we know that the length of the $i$th stretch is
\[
    x_{i+1} - x_i = \pm \left( F_i(\phi(x_i)) + F_i(\phi(x_{i+1})) \right).
\]

@@

This from ``NEQwiki, the nonlinear equations encyclopedia'', \url{http://www.primat.mephi.ru/wiki/ow.asp?Korteweg-de_Vries_equation}.
Recall that $V_i(\phi) = (1/2) s_i \phi^2(1-\phi/3) + K_i$, with $K_i = V_i(0)$ chosen to match $V$ at the boundaries.
Note that if we rescale space by $\sqrt{s}/\sigma$,
then $s$ and $\sigma$ drop from the equation.
We want to integrate
\[
    \int \frac{ d\phi }{ \sqrt{-V(\phi)} } = 
        \sqrt{ 2 } \int \frac{ d\phi }{ \sqrt{ \phi^2 (1-\phi/3) + C } } ,
\]
with $C=2K$.
Let $\phi^2(1-\phi/3)+C = (\alpha-\phi)(\phi-\beta)(\phi-\gamma)$,
and change variables first to $y^2=(\alpha-\phi)$, 
and then to $x = y/\sqrt{\alpha-\beta}$, so that
\begin{align*}
    \frac{ d\phi }{ \sqrt{ \phi^2 (1-\phi/3) + C } } 
        = \frac{ - 2 dy }{ \sqrt{ (\alpha-\beta-y^2) (\alpha-\gamma-y^2) } } \\
        = \frac{ - 2 dx }{ \sqrt{\alpha-\gamma} \sqrt{ (1-x^2) (1-S^2 x^2) } } ,
\end{align*}
with $S^2 = (\alpha-\beta)/(\alpha-\gamma)$.
Now Jacobi's ``incomplete elliptic integral of the first kind'' is
\[
    F(x;k) = \int_0^x \frac{dt}{\sqrt{ (1-t^2)(1-k^2t^2) }} ,
\]
and the Jacobian elliptic function $\sn(x;k)$ is the inverse: $F(\sn(x;k);k) = x$.

@@

Presumably, this gets the same answer as Monty, who finds that in his solvable cases,
for $x$ in the last interval (which ends in $\infty$), that
\begin{align*}
    \phi(x) &= \frac{3}{2}\left( 1 - \tanh^2 ( x \sqrt{s} / \sigma + C ) \right) \\
        &\simeq C' \exp\left( - \frac{ x \sqrt{s} }{ \sigma } \right) .
\end{align*}
Indeed, as $\phi \to 0$, we have approximately $\phi' = \frac{ \sqrt{s} }{ \sigma } \phi$.

>>>>

Let $p(t,x)$ be the proportion of the allele under spatially varying selection $s(x)$ in one dimension,
so that 
\[
\partial_t p(t,x) = \frac{1}{2} \sigma^2 \partial_x^2 p(t,x) + s(x) p(t,x) (1-p(t,x)) .
\]
The stable distribution $\phi(x) = \lim_{t\to\infty} p(t,x)$ then solves
\begin{align} \label{eqn:definingphi}
    \partial_x^2 \phi(x) = - 2 s(x) \phi(x) (1-\phi(x)) /\sigma^2,
\end{align}
with appropriate boundary conditions.
First rescale space by $\sigma/\sqrt{2}$ so the $\sigma^2/2$ term dissappears.
If we now assume that $s(x)$ is piecewise constant,
$s(x) = s_i$ for $x \in [x_i,x_{i+1})$, with $x_0=-\infty$ and $x_{n+2}=\infty$,
then the equation is integrable: if we multiply through by $2\partial_x \phi(x)$ and integrate, then we get that
\begin{align} \label{eqn:conservation}
    ( \partial_x \phi(x) )^2  &= - \int^{x} 2 s(x) \phi(x) (1-\phi(x)) \partial_x \phi(x) dx \\
        &= - s_i \phi(x)^2 \left( 1 - \frac{2}{3} \phi(x) \right) - K_i \quad \mbox{for } x \in [x_i,x_{i+1}) \\
        &:= - V_i(\phi(x)) \quad \mbox{for } x \in [x_i,x_{i+1}) ,
\end{align}
where $K_i$ is a constant of integration and we define
\[
        V_i(\phi) = s_i \phi^2 \left( 1 - \frac{2}{3} \phi \right) + K_i .
\]
Note that $V_i'(0)=V_i'(1)=0$, that $V_i(0)=K_i$ and $V_i(1) = K_i+s_i/3$.  
We will always have that $V(\phi) \le 0$.
(We have then that $\phi'' = - \partial_\phi V(\phi)$, the equation of motion of a particle in potential $V$.
Also note that this implies ``conservation of energy'', i.e.\ $( \partial_x \phi(x) )^2 + V(\phi(x))$ is constant.)
Rearranging, we get that $dx = d\phi / \sqrt{-V(\phi)}$, so where $\phi(x)$ is monotone, the inverse is
\[
    x(\phi) = x(\phi_0) \pm \int_{\phi_0}^\phi \frac{ d\psi }{ \sqrt{ -V(\psi) } } .
\]
For each $i$ then define the elliptic function
\[
    F_i(\phi) = \int_{\phi_i^*}^\phi \frac{ d\psi }{ \sqrt{ -V(\psi) } } ,
\]
where take the positive branch of the square root, and $\phi_i^*$ will be chosen later.
Then we have that
$x(\phi) - x(\phi_0) = \pm( F(\phi) - F(\phi_0))$,
or for an appropriate $x_0$,
\[
    \phi(x) = F^{-1}\left( F(\phi(x_0)) \pm (x - x_0) \right).
\]

@@

We also require that $\phi(x)$ and $\phi'(x)$ are continuous.
Continuity of $\phi'(x)$ is equivalent to $V_i(x_{i+1}) = V_{i+1}(x_{i+1})$,
which we can rearrange to find an equation for $K_{i+1}$ in terms of $K_i$ and $\phi(x_{i+1})$:
\[
    K_{i+1} - K_i = (s_i - s_{i+1}) \phi(x_{i+1})^2 (1-2\phi(x_{i+1})/3) .
\]
What about $\phi(x_i)$?  Well, if $\phi(x)$ is monotone on $[x_i,x_{i+1})$ then we can wolog take $\phi_i^*=0$ or $1$ depending on the sign of $s_i$.
Otherwise, let $\phi_i^*$ be the (unique) root of $V_i$ in $[x_i,x_{i+1})$, so $V_i(\phi_i^*)=0$.
In this case, $\phi_i^*$ is the maximum or minimum of $\phi$ in the interval:
if $s>0$, then $\phi_i = \max\{ \phi(x) : x \in [x_i,x_{i+1})\}$. 
Recall we defined $F_i$ using $\phi_i^*$; now using the fact that $\phi$ is monotone with the opposite sign on either side of $\phi_i^*$,
$x_{i+1} - x_i = F(\phi(x_{i+1})) - F(\phi_i^*) + F(\phi(x_{i})) - F(\phi_i^*)$,
and that $F(\phi_i^*) = 0$,
we know that the length of the $i$th stretch is
\[
    x_{i+1} - x_i = \pm \left( F_i(\phi(x_i)) + F_i(\phi(x_{i+1})) \right).
\]

@@

This from ``NEQwiki, the nonlinear equations encyclopedia'', \url{http://www.primat.mephi.ru/wiki/ow.asp?Korteweg-de_Vries_equation}.
Recall that $V_i(\phi) =  s_i \phi^2(1-2\phi/3) + K_i$, with $K_i = V_i(0)$ chosen to match $V$ at the boundaries.
Note that if we rescale space by $\sqrt{s}/\sigma$,
then $s$ and $\sigma$ drop from the equation.
We want to integrate
\[
    \int \frac{ d\phi }{ \sqrt{-V(\phi)} } = 
         \int \frac{ d\phi }{ \sqrt{ \phi^2 (1-2\phi/3) + K } } .
\]
Let $\phi^2(1-2\phi/3)+K = (\alpha-\phi)(\phi-\beta)(\phi-\gamma)$,
and change variables first to $y^2=(\alpha-\phi)$, 
and then to $x = y/\sqrt{\alpha-\beta}$, so that
\begin{align*}
    \frac{ d\phi }{ \sqrt{ \phi^2 (1-2\phi/3) + K } } 
        = \frac{ - 2 dy }{ \sqrt{ (\alpha-\beta-y^2) (\alpha-\gamma-y^2) } } \\
        = \frac{ - 2 dx }{ \sqrt{\alpha-\gamma} \sqrt{ (1-x^2) (1-S^2 x^2) } } ,
\end{align*}
with $S^2 = (\alpha-\beta)/(\alpha-\gamma)$.
Now Jacobi's ``incomplete elliptic integral of the first kind'' is
\[
    F(x;k) = \int_0^x \frac{dt}{\sqrt{ (1-t^2)(1-k^2t^2) }} ,
\]
and the Jacobian elliptic function $\sn(x;k)$ is the inverse: $F(\sn(x;k);k) = x$.

@@

Presumably, this gets the same answer as Monty, who finds that in his solvable cases,
for $x$ in the last interval (which ends in $\infty$), that
\begin{align*}
    \phi(x) &= \frac{3}{2}\left( 1 - \tanh^2 ( x \sqrt{2s} / \sigma + C ) \right) \\
        &\simeq C' \exp\left( - \frac{ 2 x \sqrt{2s} }{ \sigma } \right) .
\end{align*}
% Indeed, as $\phi \to 0$, we have approximately $\phi' = \frac{ \sqrt{s} }{ \sigma } \phi$.  % huh????
